Visualizing Independence Using Extended Association Plots

### 2005-07-20 trunk/vcd/paper/JCGS.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@402 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

<<preliminaries,echo=FALSE,results=hide>>=
set.seed(12345)
library(vcd)

@@

data(Employment)
emp <- aperm(Employment)
emp1 <- co_table(Employment, "LayoffCause")[["Replaced"]]
@

<<auxfunctions,echo=FALSE,results=hide>>=
## pmos.Friendly <- function(x, ...) grid.mosaicplot(x, panel = TRUE, legend = FALSE,
##   labels = FALSE, space = 0.02, gp = gp.Friendly, ...)
## pmos.max <- function(x, ...) grid.mosaicplot(x, panel = TRUE, legend = FALSE,
##   labels = FALSE, space = 0.02, gp = gp.max, ...)

hue.slice <- function(hue, grid.n = 100, type = c("HCL", "HSV"), plot = TRUE, fixup = FALSE)
{
  type <- match.arg(type)
  if(type == "HCL") {
    chroma = seq(0, 100, length = grid.n)
    luminance = seq(0, 100, length = grid.n)
    nc <- length(chroma)
    nl <- length(luminance)
    color.slice <- outer(chroma, luminance, function(y, x) hcl(hue, x, y, fixup = fixup))
    xlab <- "chroma"
    ylab <- "luminance"
    main <- paste("hue =", round(hue, digits = 0))
  } else {
    chroma = seq(0, 1, length = grid.n)
    luminance = seq(0, 1, length = grid.n)
    nc <- length(chroma)
    nl <- length(luminance)
    color.slice <- outer(chroma, luminance, function(y, x) hsv(hue, x, y))
    xlab <- "saturation"
    ylab <- "value"
    main <- paste("hue =", round(hue, digits = 3))
  }
  if(plot) {
    plot(0.5, 0.5, xlim = range(chroma), ylim = range(luminance), type = "n", axes = FALSE,
         xlab = xlab, ylab = ylab, yaxs = "i", xaxs = "i", main = main)
    for(i in 1:(nc-1)) {
      rect(chroma[i], luminance[-nl], chroma[i] + 100/(nc-1), luminance[-1], border = color.slice[,i+1], col = color.slice[,i+1])
    }
    axis(1)
    axis(2)
    box()
  }
  colnames(color.slice) <- chroma
  rownames(color.slice) <- luminance
  attr(color.slice, "type") <- type
  class(color.slice) <- "slice"
  invisible(color.slice)
}

@@

\section{Visualizations} \label{sec:visualization}

@@

@@

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[htbp]
\begin{center}
<<goals-assoc-classic,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
assoc(emp1)
@
<<goals-mosaic-classic,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(emp1, split_vertical = TRUE)
@
\caption{Classic association and mosaic plot for the ``Replaced'' subset of the Employment data.}
\label{fig:classic}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
<<art-Friendly,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(art, gp = shading_Friendly)
@
<<art-max,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(art, gp = shading_max)
@
\caption{Mosaic plots for the Arthritis data.}
\label{fig:art}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
<<Friendly-HSV,fig=TRUE,echo=FALSE,height=7,width=8>>=
## generate colors
hue23 <- hue.slice(2/3, grid.n = 101, plot = FALSE, type = "HSV")
hue0 <- hue.slice(0, grid.n = 101, plot = FALSE, type = "HSV")
mysaturation <- as.numeric(colnames(hue23))
value <- as.numeric(rownames(hue23))

@@

<<HCL-space,fig=TRUE,echo=FALSE,height=7,width=8>>=
## generate colors
hue260 <- hue.slice(260, grid.n = 101, plot = FALSE)
hue360 <- hue.slice(360, grid.n = 101, plot = FALSE)
mychroma <- as.numeric(colnames(hue260))
luminance <- as.numeric(rownames(hue260))

@@

lines(c(-18.75, -11.25), c(52.5, 67.5), lty = 2)
lines(c(-8.75, -1.25), c(72.5, 87.5), lty = 2)
lines(c(18.75, 11.75), c(52.5, 67.5), lty = 2)
lines(c(8.75, 1.25), c(72.5, 87.5), lty = 2)
@
\caption{Extended shading in HSV and HCL space.}
\label{fig:ExtHSVHCL}
\end{center}
\end{figure}

\setkeys{Gin}{width=\textwidth}
\begin{figure}[htbp]
\begin{center}
<<UCB-mosaic1,fig=TRUE,echo=FALSE,height=6,width=8>>=
## fm <- loglm(~(EmploymentStatus+EmploymentLength)*LayoffCause, data = emp, fitted = TRUE)
## my.gp <- gp.HCLshading(emp, residuals(fm, type = "pearson"))
## my.gp$p.value <- summary(fm)$tests[2,3]
## my.gp$lty <- c(1,1)
## grid.mosaicplot(emp, gp = my.gp, margin = ~(EmploymentStatus+EmploymentLength)*LayoffCause,
##   space = c(0.1, 0.01, 0), dir = "v")
mosaic(emp, expected = ~(EmploymentStatus+EmploymentLength)*LayoffCause)
@
\caption{Employment data with mosaic display.}
\label{fig:emp-mosaic1}
\end{center}
\end{figure}

@@

\setkeys{Gin}{width=0.8\textwidth}
\begin{figure}[htbp]
\begin{center}
<<Friendly-hue,fig=TRUE,echo=FALSE,height=4.8,width=8>>
plot(0.5, 0.5, xlim = c(0, 1), ylim = c(0, 1), type = "n", axes = FALSE, xlab = "", ylab = "", yaxs = "i")
dummy <- 0:100/100
dummy.col <- hsv(dummy[-101] + 1/200, 1, 1)
rect(dummy[-101], 0.6, dummy[-1], 1, border = dummy.col, col = dummy.col)
rect(0, 0.6, 1, 1)
axis(3)
mtext("hue", line = 2, cex = 1.5)
text(1, 0.45, "saturation = 1\nvalue = 1", pos = 2, cex = 1.2)
lines(c(0.1, 0), c(0.3, 0.6), lty = 2)
lines(c(2/3, 2/3), c(0.3, 0.6), lty = 2)
rect(0, 0.1, 0.2, 0.3, col = hsv(0, 1, 1))
rect(2/3 - 0.1, 0.1, 2/3 + 0.1, 0.3, col = hsv(2/3, 1, 1))
text(0.1, 0.1, expression(r[ij] < 0), pos = 1, cex = 1.5)
text(2/3, 0.1, expression(r[ij] > 0), pos = 1, cex = 1.5)
@

@@

\begin{figure}[htbp]
\begin{center}
<<emp1-Friendly,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>
mosaic(emp1, gp = shading_Friendly)
@
<<emp1-max,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>
mosaic(emp1, gp = shading_max)
@
\caption{Mosaic plots for the ``Replaced'' subset of the Employment data.}
\label{fig:emp1}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
<<UCB-mosaic2,fig=TRUE,echo=FALSE,height=4,width=8>>
tabplot(emp, panel = pmos.max)
@
\caption{Employment data with conditioned mosaic display.}
\label{fig:emp-mosaic2}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
<<UCB-assoc,fig=TRUE,echo=FALSE,height=4,width=8>>
tabplot(emp)
@
\caption{Employment data with conditioned association plot.}
\label{fig:emp-assoc}
\end{center}
\end{figure}

>>>>

<<load_data,echo=FALSE,results=hide>>=
library(vcd)

@@

data(UCBAdmissions)
ucb <- aperm(UCBAdmissions)
@

<<auxiliary_functions,echo=FALSE,results=hide>>=
hue.slice <- function(hue, grid.n = 100, type = c("HCL", "HSV"), plot = TRUE, fixup = FALSE)
{
  type <- match.arg(type)
  if(type == "HCL") {
    chroma = seq(0, 100, length = grid.n)
    luminance = seq(0, 100, length = grid.n)
    nc <- length(chroma)
    nl <- length(luminance)
    color.slice <- outer(chroma, luminance, function(y, x) hcl(hue, x, y, fixup = fixup))
    xlab <- "chroma"
    ylab <- "luminance"
    main <- paste("hue =", round(hue, digits = 0))
  } else {
    chroma = seq(0, 1, length = grid.n)
    luminance = seq(0, 1, length = grid.n)
    nc <- length(chroma)
    nl <- length(luminance)
    color.slice <- outer(chroma, luminance, function(y, x) hsv(hue, x, y))
    xlab <- "saturation"
    ylab <- "value"
    main <- paste("hue =", round(hue, digits = 3))
  }
  if(plot) {
    plot(0.5, 0.5, xlim = range(chroma), ylim = range(luminance), type = "n", axes = FALSE,
         xlab = xlab, ylab = ylab, yaxs = "i", xaxs = "i", main = main)
    for(i in 1:(nc-1)) {
      rect(chroma[i], luminance[-nl], chroma[i] + 100/(nc-1), luminance[-1], border = color.slice[,i+1], col = color.slice[,i+1])
    }
    axis(1)
    axis(2)
    box()
  }
  colnames(color.slice) <- chroma
  rownames(color.slice) <- luminance
  attr(color.slice, "type") <- type
  class(color.slice) <- "slice"
  invisible(color.slice)
}

@@

\subsection{Visualizations} \label{sec:visualization}

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[htbp]
\begin{center}
<<classic_plots,fig=TRUE,echo=FALSE,height=6,width=11,results=hide>>=
grid.newpage()
pushViewport(viewport(layout = grid.layout(1, 2)))
pushViewport(viewport(layout.pos.col=1, layout.pos.row=1))
mosaic(art, newpage = FALSE, margins = c(2.5, 4, 2.5, 3))
popViewport()
pushViewport(viewport(layout.pos.col=2, layout.pos.row=1))
assoc(art, newpage = FALSE, margins = c(5, 2, 5, 4))
popViewport(2)
@
\caption{Classic mosaic and association plot for the Arthritis data.}
\label{fig:classic}
\end{center}
\end{figure}

@@

\setkeys{Gin}{width=.8\textwidth}
\begin{figure}[p]
\begin{center}
<<shading_HSV,fig=TRUE,echo=FALSE,height=7,width=8>>=
## generate colors
hue23 <- hue.slice(2/3, grid.n = 101, plot = FALSE, type = "HSV")
hue0 <- hue.slice(0, grid.n = 101, plot = FALSE, type = "HSV")
mysaturation <- as.numeric(colnames(hue23))
value <- as.numeric(rownames(hue23))

@@

<<shading_HCL,fig=TRUE,echo=FALSE,height=7,width=8>>=
## generate colors
hue260 <- hue.slice(260, grid.n = 101, plot = FALSE)
hue360 <- hue.slice(360, grid.n = 101, plot = FALSE)
mychroma <- as.numeric(colnames(hue260))
luminance <- as.numeric(rownames(hue260))

@@

lines(c(-18.75, -11.25), c(52.5, 67.5), lty = 2)
lines(c(-8.75, -1.25), c(72.5, 87.5), lty = 2)
lines(c(18.75, 11.75), c(52.5, 67.5), lty = 2)
lines(c(8.75, 1.25), c(72.5, 87.5), lty = 2)
@
\caption{Extended shading in HSV and HCL space.}
\label{fig:shadingHSVHCL}
\end{center}
\end{figure}

\setkeys{Gin}{width=\textwidth}
\begin{figure}[htbp]
\begin{center}
<<shadings,fig=TRUE,echo=FALSE,height=6,width=11,results=hide>>=
grid.newpage()
pushViewport(viewport(layout = grid.layout(1, 2)))
pushViewport(viewport(layout.pos.col=1, layout.pos.row=1))
mosaic(art, newpage = FALSE, gp = shading_Friendly, margins = c(2.5, 4, 2.5, 3))
popViewport()
pushViewport(viewport(layout.pos.col=2, layout.pos.row=1))
set.seed(1234)
mosaic(art, newpage = FALSE, gp = shading_max, margins = c(2.5, 3, 2.5, 4))
popViewport(2)
@
\caption{Mosaic plot for the Arthritis data with Friendly shading (left) and extended shading (right).}
\label{fig:shadings}
\end{center}
\end{figure}

\setkeys{Gin}{width=\textwidth}
\begin{figure}[htbp]
\begin{center}
<<conditional_assoc,fig=TRUE,echo=FALSE,height=6,width=9>>=
set.seed(1234)
cotabplot(~ Admit + Gender | Dept, data = ucb, panel = cotab_coindep, type = "assoc")
@
\caption{Conditional association plot for UCB admissions.}
\label{fig:emp-mosaic1}
\end{center}
\end{figure}

@@

%% for reproducibility, print also

<<graphics>>
## load packages and data
library(vcd)
data(Arthritis)
art <- xtabs(~ Treatment + Improved, data = Arthritis, subset = Sex == "Female")
data(UCBAdmissions)
ucb <- aperm(UCBAdmissions)

## plain classic plots
mosaic(art)
assoc(art)

## new max shading with HCL colors
mosaic(art, gp = shading_Friendly)
set.seed(1234)
mosaic(art, gp = shading_max)

## conditional independence
set.seed(1234)
cotabplot(~ Admit + Gender | Dept, data = ucb, panel = cotab_coindep, type = "assoc")
## which uses the double max functional
set.seed(1234)
coindep_test(ucb, "Dept")

## alternatively use Peason functional
## via simulation of conditional distribution:
set.seed(1234)
coindep_test(ucb, "Dept", indepfun = function(x) sum(x^2), aggfun = sum)
## via asymptotic distribution:
fm <- loglm(~ Dept * (Gender + Admit), data = ucb)
fm
plot(fm, labeling_args = list(abbreviate = c(Admit = 2)))
@

%% deprecated stuff

%% employment data
<<employment>>
## data
data(Employment)
emp <- aperm(Employment)
emp1 <- co_table(Employment, "LayoffCause")[["Replaced"]]

## plain
assoc(emp1)
mosaic(emp1)

## new shadings
mosaic(emp1, gp = shading_Friendly)
mosaic(emp1, gp = shading_max)

## conditional
mosaic(emp, expected = ~(EmploymentStatus+EmploymentLength)*LayoffCause)
cotabplot(emp, "LayoffCause", split_vertical = TRUE)
cotabplot(emp, panel = cotab_coindep, type = "assoc")
cotabplot(emp, "LayoffCause", panel = cotab_coindep, type = "assoc", split_vertical = TRUE)
@

%% Friendly shading in detail
\setkeys{Gin}{width=0.8\textwidth}
\begin{figure}[htbp]
\begin{center}
<<Friendly-hue,fig=TRUE,echo=FALSE,height=4.8,width=8>>
plot(0.5, 0.5, xlim = c(0, 1), ylim = c(0, 1), type = "n", axes = FALSE, xlab = "", ylab = "", yaxs = "i")
dummy <- 0:100/100
dummy.col <- hsv(dummy[-101] + 1/200, 1, 1)
rect(dummy[-101], 0.6, dummy[-1], 1, border = dummy.col, col = dummy.col)
rect(0, 0.6, 1, 1)
axis(3)
mtext("hue", line = 2, cex = 1.5)
text(1, 0.45, "saturation = 1\nvalue = 1", pos = 2, cex = 1.2)
lines(c(0.1, 0), c(0.3, 0.6), lty = 2)
lines(c(2/3, 2/3), c(0.3, 0.6), lty = 2)
rect(0, 0.1, 0.2, 0.3, col = hsv(0, 1, 1))
rect(2/3 - 0.1, 0.1, 2/3 + 0.1, 0.3, col = hsv(2/3, 1, 1))
text(0.1, 0.1, expression(r[ij] < 0), pos = 1, cex = 1.5)
text(2/3, 0.1, expression(r[ij] > 0), pos = 1, cex = 1.5)
@

@@
